### 2025-07-10 paper/quarto_aaai/paper.tex

% REMOVE THIS: bibentry
% This is only needed to show inline citations in the guidelines document. You should not need it and can safely delete it.
% \usepackage{bibentry}
% END REMOVE bibentry
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
% For mathematical environments, you may reduce fontsize but not below 6.5 point.
\DeclareMathSizes{10}{9}{7}{6.5}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  To what extent does the counterfactual training objective as it is
  defined in Equation~\ref{eq-obj} induce models to learn plausible
  explanations?
\item
  To what extent does CT produce more favorable AR outcomes in the
  presence of actionability constraints?
\item
  What are the effects of hyperparameter selection on CT?
\end{enumerate}

>>>>

% REMOVE THIS: bibentry
% This is only needed to show inline citations in the guidelines document. You should not need it and can safely delete it.
\usepackage{bibentry}
% END REMOVE bibentry
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\usepackage{placeins}
\usepackage{siunitx}
\sisetup{uncertainty-mode = separate}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  To what extent does the CT objective in Equation 1 induce models to
  learn plausible explanations?
\item
  To what extent does the CT objective produce more favorable
  algorithmic recourse outcomes in the presence of actionability
  constraints?
\item
  What are the effects of hyperparameter selection wrt. the CT
  objective?
\end{enumerate}

### 2025-07-10 paper/sections/conclusion.qmd

<!-- # Conclusion {#sec-conclusion} -->
To conclude, state-of-the-art machine learning models are prone to learning complex representations that cannot be interpreted by humans and existing post-hoc explainability approaches cannot guarantee that the explanations agree with the model's learned representation of data. As a step towards addressing this challenge, we introduced counterfactual training, a novel training regime that incentivizes highly-explainable models. Our approach leads to explanations that are both plausible---compliant with the underlying data-generating process---and actionable---compliant with user-specified mutability constraints---and thus meaningful to their recipients. Through extensive experiments we demonstrate that CT satisfies its objective while promoting robustness and preserving the predictive performance of the models. It can also be used to fine-tune conventionally-trained models and achieve similar gains. Finally, this work showcases that it is practical to improve models *and* their explanations at the same time.

>>>>

To conclude, state-of-the-art machine learning models are prone to learning complex representations that cannot be interpreted by humans. Existing explainability solutions cannot guarantee that explanations agree with the model's learned representation of data. As a step towards addressing this challenge, we introduced counterfactual training, a novel training regime that incentivizes highly-explainable models. Our approach leads to explanations that are both plausible (compliant with the underlying data-generating process) and actionable (compliant with user-specified mutability constraints), and thus meaningful to their recipients. Through extensive experiments, we demonstrated that CT satisfies its objective while promoting robustness and preserving the predictive performance of models. It can also be used to fine-tune conventionally-trained models and achieve similar gains. Thus, our work showcases that it is practical to simultaneously improve models and their explanations.

### 2025-07-10 paper/sections/discussion.qmd

*CT increases the training time of models.* CT can be more time-consuming than conventional training regimes. Higher numbers of CEs per iteration positively impact the quality of solutions, but they also increase the amount of computations. In our experiments, grids with 270 settings took up to four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs (see supplementary appendix for details). Three factors attenuate this effect: (1) CT inherently yields itself to parallel execution; (2) it amortizes the cost of CEs for the training samples; (3) it retains its value when used to fine-tune conventionally-trained models.

@@

>>>>

@@

*CT increases the training time of models.* Just like adversarial training, CT is more resource-intensive than conventional training regimes. Higher numbers of CEs per iteration improve the quality of learned representations, but they also increase the number of computations. As our codebase is not performance optimized, grids of 270 settings took up to four hours for more demanding datasets in our experiments, when ran on 34 2GB CPUs (see supplementary appendix). Other than optimization, three factors mitigate this effect: (1) CT yields itself to parallel execution; (2) it amortizes the cost of CEs for the training samples; (3) it can be used to fine-tune conventionally-trained models.

### 2025-07-10 paper/sections/experiments.qmd

1. To what extent does the counterfactual training objective as it is defined in @eq-obj induce models to learn plausible explanations?
2. To what extent does CT produce more favorable AR outcomes in the presence of actionability constraints?
3. What are the effects of hyperparameter selection on CT?

>>>>

1. To what extent does the CT objective in Equation 1 induce models to learn plausible explanations?
2. To what extent does the CT objective produce more favorable algorithmic recourse outcomes in the presence of actionability constraints?
3. What are the effects of hyperparameter selection wrt. the CT objective?

### 2025-07-10 paper/sections/method.qmd

## Our Proposed Objective

Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote a counterfactual generated through gradient descent over $T$ iterations as originally proposed by @wachter2017counterfactual. In broad terms, searching for CEs using gradient descent entails optimizing some form of an objective that balances (1) the classification loss $\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y})$, and (2) one or more penalty terms $\lambda_{i}\text{cost}_{i}(\cdot)$. The exact specification of these penalties induces various properties in the counterfactual outcomes, and tends to be the key feature that distinguishes various gradient-based "generators" or "explainers" (CE algorithms) from the literature [@altmeyer2023explaining].

>>>>

## Proposed Objective

Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote a counterfactual generated through gradient descent over $T$ iterations as originally proposed by @wachter2017counterfactual. In broad terms, searching for CEs using gradient descent entails optimizing some form of an objective that balances (1) the classification loss $\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y})$, and (2) one or more penalty terms $\lambda_{i}\text{cost}_{i}(\cdot)$. The exact specification of these penalties induces various properties in the counterfactual outcomes, and tends to be the key feature that distinguishes various gradient-based "generators" or "explainers" (CE algorithms) from the literature [@altmeyer2023explaining], including all generators used in our experiments. We refer the reader to the supplementary appendix for details.
